Wastewater Surveillance for SARS-CoV-2 Data Analysis Project

### 2023-04-24 code/processing_code/processing.qmd
fixing issues

recovery_output \<- read_csv(here("data/raw_data/recovery_data.csv"))

recovery_input \<- read_csv(here("data/raw_data/calfguard.csv"))

```{r}
here()
# Load CFX N1 data
n1 <- read_csv(here("data/raw_data/cfx_n1.csv"))

# Load CFX N2 data
n2 <- read_csv(here("data/raw_data/cfx_n2.csv"))

@@

### Catchment-level COVID-19 Case Data

`covid_case_wrf` \<- Count of confirmed cases of COVID-19 on a given day from each WRF catchment zone in ACC.

```{r}
#I don't have access to this data yet, waiting on DPH partner to send.
```

@@

Currently, I have 7 unique sets of data. In order to do analysis on the data down the road, I want to figure out how to best combine these data sets.

@@

1.  CFX N1 results
2.  CFX N2 results
3.  Plant flow and total suspended solids
4.  County-level COVID symptoms
5.  County-level COVID confirmed cases
6.  County-level COVID tests administered
7.  Catchment-level COVID confirmed cases
8.  CFX recovery input
9.  CFX recovery output

@@

@@

copy_num_uL_rxn = as.numeric(10\^((ct-36.046)/-3.5293)), #use standard curve slope here

@@

```{r}
#N1 data here
n1 <- n1 %>% mutate(ct = as.numeric(ct), 
              copy_num_uL_rxn = as.numeric(10^((ct-36.046)/-3.5293)), #use standard curve slope here
              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004), #replace NAs with limit of detection (0.004 copies per uL of the reaction)
              run_date = as.Date(run_date))

#N2 data here (because they have different standard curve equations)
n2 <- n2 %>% mutate(ct = as.numeric(ct),
              copy_num_uL_rxn = as.numeric(10^((ct-37.731)/-3.2505)),
              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004),
              run_date = as.Date(run_date))

@@

@@

wbe_year2 <- bind_rows(n1, n2) %>% #combine dataframes with same variables
  mutate(
    date=as.Date(sample_date, format = "%d-%b-%y"), #change to date format
    facility=substr(sample_id, 1,2), #first two letters in sample_id is facility ID
    biological_replicate=substr(sample_id, nchar(sample_id), nchar(sample_id)), #last number in sample_id is the replicate number
    ct=as.numeric(ifelse(ct=="Undetermined", NA, ct)) #set NAs
  ) %>% 
  arrange(date, facility, target, biological_replicate) %>% 
  select(date, facility, target, biological_replicate, collection_num, ct, copy_num_uL_rxn) %>%
  mutate(facility = as.factor(facility), 
         facility = recode(facility, NO = "A", MI = "B", CC = "C"), 
         facility = ordered(facility, levels = c("A", "B", "C")), #set facility to factor and then set factor levels
         copy_num_L = copy_num_uL_rxn*20/5*60/280*1000*1000) %>% 
  filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))

@@

```{r}
n1_n2_plant <- left_join(wbe_year2_avg, plant_clean, by = c("date", "facility")) %>% 
  mutate(total_copies = avg_copy_L * influent_flow_L)

@@

```{r}
all_data <- full_join(n1_n2_plant, covid, by = "date") %>% 
   filter(between(date, as.Date("2021-07-06") , as.Date("2022-12-21")))

@@

### Combining recovery data sets

```{r}
recovery_calc <- left_join(recovery_output_clean, recovery_input_clean, by = "cg_num") %>% 
  mutate(percent_recovery = (copies_ul_sample/copies_ul_input)*100) %>% 
  select(facility, collection_num, percent_recovery)

saveRDS(recovery_calc, here("data/processed_data/recovery_percentages.rds"))
```

@@

@@

`pos_rate_tr` will calculate the percento of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)

@@

```{r}
wbe_br <- wbe_tr %>%
  group_by(date, facility, target) %>% 
  dplyr::summarise(
    n.bio = n(),
    n.bio.non.miss = sum(!is.na(ct.mean)),
    n.bio.miss = sum(is.na(ct.mean)),
    n.total = sum(n), 
    n.total.miss = sum(n.miss), 
    bio.ct.mean = mean(ct.mean, na.rm = T), 
    bio.ct.sd = sd(ct.mean, na.rm=T), 
    pos_rate_tr = mean(percent_pos),
    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
  ) %>%
  mutate(percent_pos_bio = (n.bio.non.miss/n.bio)*100) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup()
wbe_br
saveRDS(wbe_br, here("data/processed_data/prop_pos_bio.rds"))
```

@@

@@

final_data
```

@@

```{r}
final_data_sum <- final_data %>% group_by(date, collection_num, target) %>% 
  summarize(sum_total_copies = sum(total_copies),
            pos_rate_tr = mean(pos_rate_tr),
            percent_pos_test = mean(percent_pos_test),
            case_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(target) %>% 
  summarize(avg_VL = mean(sum_total_copies),
            pos_rate_tr = mean(pos_rate_tr),
            percent_pos_test = mean(percent_pos_test),
            case_7dma = mean(case_7dma)) %>% 
  ungroup() %>% 
  mutate(log10_VL = log10(avg_VL))

saveRDS(final_data_sum, here("data/processed_data/final_sum_combined_data.RDS"))
```

>>>>

```{r, message=FALSE}
here()
# Load CFX N1 data
n1_1 <- read_csv(here("data/raw_data/n1_data_year1.csv")) #year 1 data
n1 <- read_csv(here("data/raw_data/cfx_n1.csv")) #year 2 data

# Load CFX N2 data
n2_1 <- read_csv(here("data/raw_data/n2_data_year1.csv")) #year 1 data
n2 <- read_csv(here("data/raw_data/cfx_n2.csv")) #year 2 data

@@

@@

Currently, I have 6 datasets. In order to do analysis on these data down the road, I want to figure out how to best combine them.

@@

1.  CFX N1 results
2.  CFX N2 results
3.  WWTP flow rate and total solids
4.  County-level COVID symptoms
5.  County-level COVID confirmed cases
6.  County-level COVID tests administered

@@

### Remove unnecessary variables

```{r}
#N1
n1_1 <- n1_1 %>% select(c(sample_date, collection_num, sample_id, target, ct))
n1 <- n1 %>% select(c(sample_date, collection_num, sample_id, target, ct)) %>% 
  filter(between(sample_date, as.Date("2021-07-06") , as.Date("2022-12-21")))

#N2
n2_1 <- n2_1 %>% select(c(sample_date, collection_num, sample_id, target, ct))
n2 <- n2 %>% select(c(sample_date, collection_num, sample_id, target, ct)) %>% 
  filter(between(sample_date, as.Date("2021-07-06") , as.Date("2022-12-21")))
```

@@

Year1:

Year 2: copy_num_uL_rxn = as.numeric(10\^((ct-36.046)/-3.5293)), #use standard curve slope here

@@

```{r}
#year 1 N1 data here
n1_1 <- n1_1 %>% mutate(ct = as.numeric(ct), 
                copy_num_uL_rxn = as.numeric(10^((ct-34.008)/-3.389)),
                copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))

#year 2 N1 data here
n1 <- n1 %>% mutate(ct = as.numeric(ct), 
              copy_num_uL_rxn = as.numeric(10^((ct-36.046)/-3.5293)), #use standard curve slope here
              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004)) #replace NAs with limit of detection (0.004 copies per uL of the reaction)

#N2 year 1 data here
n2_1 <-n2_1 %>% mutate(ct = as.numeric(ct),
                copy_num_uL_rxn = as.numeric(10^((ct-32.416)/-3.3084)),
                copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))

#N2 year 2 data here
n2 <- n2 %>% mutate(ct = as.numeric(ct),
              copy_num_uL_rxn = as.numeric(10^((ct-37.731)/-3.2505)),
              copy_num_uL_rxn = replace_na(copy_num_uL_rxn, 0.004))

@@

### Combining year 1 and 2 data

```{r}
#N1
n1 <- bind_rows(n1_1, n1)
tibble(n1)

#N2
n2 <- bind_rows(n2_1, n2)
tibble(n2)
```

@@

wbe_year2 <- bind_rows(n1, n2) %>% #combine dataframes with same variables
  mutate(
    date=as.Date(sample_date, format = "%d-%b-%y"), #change to date format
    facility=substr(sample_id, 1,2), #first two letters in sample_id is facility ID
    biological_replicate=substr(sample_id, nchar(sample_id), nchar(sample_id)), #last number in sample_id is the replicate number
    ct=as.numeric(ifelse(ct=="Undetermined", NA, ct)) #set NAs
  ) %>% 
  arrange(date, facility, target, biological_replicate) %>% 
  select(date, facility, target, biological_replicate, collection_num, ct, copy_num_uL_rxn) %>%
  mutate(facility = as.factor(facility), 
         facility = recode(facility, NO = "A", MI = "B", CC = "C"), 
         facility = ordered(facility, levels = c("A", "B", "C")), #set facility to factor and then set factor levels
         copy_num_L = copy_num_uL_rxn*20/5*60/280*1000*1000)

@@

```{r}
n1_n2_plant <- left_join(wbe_year2, plant_clean, by = c("date", "facility")) %>% 
  mutate(total_copies = copy_num_L * influent_flow_L)

@@

```{r}
all_data <- full_join(n1_n2_plant, covid, by = "date") %>% 
   filter(between(date, as.Date("2020-06-30") , as.Date("2022-12-21")))

@@

@@

#### Percent positive technical reps for N1 only

```{r}
wbe_tr_n1 <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
  filter(target == "N1") %>% 
  dplyr::summarise(
    n=n(), 
    n.miss=sum(is.na(ct)), 
    ct.mean=mean(ct,na.rm=T), 
    ct.sd=sd(ct,na.rm=T)
  ) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(n.pos = n - n.miss) %>% 
  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
  mutate(percent_pos = (n.pos/n)*100)
wbe_tr_n1
```

#### Percent positive N2 only

```{r}
wbe_tr_n2 <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
  filter(target == "N2") %>% 
  dplyr::summarise(
    n=n(), 
    n.miss=sum(is.na(ct)), 
    ct.mean=mean(ct,na.rm=T), 
    ct.sd=sd(ct,na.rm=T)
  ) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(n.pos = n - n.miss) %>% 
  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
  mutate(percent_pos = (n.pos/n)*100)
wbe_tr_n2
```

@@

`pos_rate_tr` will calculate the percent of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)

@@

```{r}
wbe_br <- wbe_tr %>%
  group_by(date, facility, target) %>% 
  dplyr::summarise(
    n.bio = n(),
    n.bio.non.miss = sum(!is.na(ct.mean)),
    n.bio.miss = sum(is.na(ct.mean)),
    n.total = sum(n), 
    n.total.miss = sum(n.miss), 
    bio.ct.mean = mean(ct.mean, na.rm = T), 
    bio.ct.sd = sd(ct.mean, na.rm=T), 
    pos_rate_tr = mean(percent_pos)) %>%
  mutate(percent_pos_bio = (n.bio.non.miss/n.bio)*100) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup()
wbe_br
saveRDS(wbe_br, here("data/processed_data/prop_pos_bio.rds"))
```

@@

### Calculate percent positive per collection date

```{r}
wbe_coll <- wbe_samp %>%
  group_by(date) %>% 
  dplyr::summarise(
    n.pos = sum(!is.na(n.pos)),
    n.miss = sum(is.na(n.miss)),
    pos_rate_tr = mean(pos_rate_tr)) %>%
   mutate(prop.pos = n.pos/(n.pos+n.miss)*100) %>% 
  #mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup()
```

@@

@@

```{r}
final_data_sum <- final_data %>% group_by(date, collection_num, target) %>% 
  summarize(sum_total_copies = sum(total_copies),
            pos_rate_tr = mean(pos_rate_tr),
            percent_pos_test = mean(percent_pos_test),
            case_7dma = mean(pcr_pos_7dma),
            pcr_pos = mean(pcr_pos),
            test_admin = mean(pcr_tests)) %>% 
  ungroup(target) %>% 
  summarize(avg_VL = mean(sum_total_copies),
            pos_rate_tr = mean(pos_rate_tr),
            percent_pos_test = mean(percent_pos_test),
            case_7dma = mean(case_7dma),
            pcr_pos = mean(pcr_pos),
            test_admin = mean(test_admin)) %>% 
  ungroup() %>% 
  mutate(log10_VL = log10(avg_VL)) %>% 
  drop_na()

table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"

saveRDS(final_data_sum, here("data/processed_data/final_sum_combined_data.RDS"))

final_data_sum

table1 <- (table1(~ log10_VL + pos_rate_tr + case_7dma + pcr_pos + percent_pos_test, data=final_data_sum)) 
saveRDS(table1, here("results/manuscript_table1.rds"))
```

### 2023-04-24 products/manuscript/Manuscript.qmd
fixing issues

---
title: "Data Analysis Project"
subtitle: ""
author: Leah Lariscy
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

*Write a summary of your project.*

@@

Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production.

Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers.

Wastewater-based detection methods have been utilized across the globe as an independent indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends \[4\]. WBE methods may also become a primary tool in understanding transmission dynamics as case under reporting increases.

@@

24-hour composite samples collected twice weekly, stored at 4C until ready for extraction. 6 extraction replicates were performed for each collection date at treatment facility.

2019-nCoV N1 and N2 primer-probe sets were used for universal detection of SARS-CoV-2.

The concentration per reaction of each gene target was determined by the corresponding standard curve, then sample concentration in copies per liter was determined by Equation 1. Total daily copies per day per WRF was determined by Equation 2.

## Questions/Hypotheses to be addressed

Is there a relationship between viral load and clinical case observations?

Is there a relationship between qPCR percent positive and clinical case observations?

What do viral load trends look like over time?

What do clinical case trends look like over time?

How do each of these temporal trends compare to one another over time?

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

@@

## Data aquisition

Wastewater SARS-CoV-2 data was obtained through weekly surveillance efforts utilizing RT-qPCR methods.

Water Reclamation Facility data was obtained through data share directly from the WRF representatives.

COVID data downloaded from Georgia DPH website.

@@

*qPCR data*

WWTP data

*DPH COVID data*

@@

Simple analyses: Wastewater and clinical data over time, linear regression model of wastewater vs clinical case data

Full analyses: Wastewater data as predictor for clinical case outcomes

Steps I took to build and tune model

@@

## Exploratory/Descriptive analysis

@@

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

>>>>

---
title: "Wastewater Surveillance for SARS-CoV-2 Data Analysis Project"
subtitle: ""
author: Leah Lariscy
date: "`r Sys.Date()`"
format:
  html:
    toc: false
    number-sections: true
    number-depth: 2
    highlight-style: github
bibliography: ../citation.bib
csl: ../apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load packages
library(here)
library(knitr)
library(tidyverse)
library(skimr)
```

# Abstract

Infection by Severe Acute Respiratory Syndrome Coronavirus 2 (SARS-CoV-2), causing COVID-19, is followed by the shedding of viral particles by multiple excretory functions, including stool and urine production. Subsequently, these viral particles can be detected in wastewater influent via RNA extraction, followed by Reverse Transcriptase Quantitative Polymerase Chain Reaction (RT-qPCR) utilizing SARS-CoV-2 specific primers. Wastewater-based detection methods have been utilized across the globe as an independent and parallel indicator of SARS-CoV-2 viral prevalence and, depending on community-specific factors, as a leading indicator of clinical case trends. WBE methods may also become a primary tool in understanding transmission dynamics as case under-reporting increases.

@@

Paragraph 1: Impact of COVID-19 on public health, clinical case under-reporting

Paragraph 2: Wastewater surveillance is possible because of excretory functions and stability of viral particles in wastewater, surveillance prevalence and correlations with clinical case data

Paragraph 3: Surveillance program background. Aims, County-level parallel analysis of wastewater surveillance data and reported clinical cases

@@

### SARS-CoV-2 wastewater surveillance data

Sample collection for wastewater surveillance began June 30, 2020 and occurred twice weekly (excluding major holidays) through December 21, 2022. Raw wastewater influent was collected from three Wastewater Treatment Plants in Athens-Clarke County and viral RNA was extracted. RT-qPCR was used to calculate the cycle threshold (Ct) value of each reaction, which is used to estimate SARS-CoV-2 copies per liter of wastewater in the data cleaning process (Section 3.2).

### Wastewater treatment plant data

Wastewater influent flow data, including millions of gallons per day (MGD) and total suspended solids (TSS), were collected for each corresponding sampling date.

### COVID-19 clinical data

COVID-19 clinical data were obtained from the Georgia Department of Public Health website [@COVID19StatusReport] which includes COVID-19 reported clinical case data for the state of Georgia. These data include reported cases, demographics, comorbidities, deaths, and more. However, not all data is necessary for analysis and will be subset to include only symptom onset, positive case, and administered test data for Clarke County. (Section 3.2).

## Hypotheses to be addressed

This study aims to conduct a parallel analysis between SARS-CoV-2 wastewater surveillance data and reported COVID-19 clinical case data in Athens-Clarke County, Georgia. It is expected that wastewater quantifications and reported case data will have a positive linear relationship. It is also hypothesized that certain wastewater-clinical relationships will be stronger than others, see Table 1 for variables of interest.

@@

## Data acquisition

### Wastewater surveillance data

24-hour composite raw wastewater influent samples were collected twice-weekly from three treatment facilities in Clarke County and stored at 4°C until ready for extraction. 280 𝜇L of wastewater in replicates (n=6) was aliquoted and direct RNA extraction (Zymo Research, R2042) was performed on the same day as sample collection.Each extraction replicate was reverse transcribed and amplified using RT-qPCR and a SARS-CoV-2 specific primer-probe panel (IDT, 10006713), done in technical replicates (n=3). As part of the qPCR panel, two gene targets on the N-gene of the SARS-CoV-2 genome were used for amplification. The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus \[\@\]. Cycle threshold (Ct) values were then calculated for each PCR reaction, which were then used to estimate SARS-CoV-2 viral load (See Section 3.2 for more detail).

### Wastewater treatment plant data

For each treatment plant and collection date, MGD and TSS was shared by plant operators directly with our team and compiled throughout the study period.

### COVID-19 clinical data

Reported case and testing data were downloaded from Georgia Department of Public Health's COVID-19 Status Report Page in January 2023.

@@

### Calculating wastewater viral load

Amplification standard curve equations where generated from serially diluted synthetic controls (Supplementary Table). This was done for N1 and N2 gene targets for both surveillance years because RT-qPCR methods were altered after the first year. Viral copies per microliter (cp/uL) of each PCR reaction was estimated by transforming the Ct value using the appropriate standard curve equation. For reactions with no detection, concentration was set to the theoretical limit of detection (0.004 cp/uL). Data sets for each standard curve was transformed separately before being combined. After combining all qPCR data, copies per liter (cp/L) of wastewater was calculated for each technical replicate using Equation 1. Average cp/L of wastewater was determined by taking the mean of all technical replicates per sample per gene target.

INSERT EQUATION 1

WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day. qPCR data was then combined with WWTP data and total viral load per sample was calculated by multiplying average copies per liter by the flow rate. Average viral load for each sample was then determined by taking the mean of the viral load for both N1 and N2 gene targets. County-level daily viral load was calculated by taking the sum of the average viral load from each of the three WWTP samples.

### Calculating wastewater qPCR assay positivity

qPCR assay positivity was determined for each collection date by finding the proportion of assays that detected SARS-CoV-2 out of the total number of assays performed. This includes all biological and technical replicate assays that were performed for a given sampling date. County-level assay positivity data was then combined with county-level viral load data.

### Adding COVID-19 clinical data

Of the downloaded Georgia DPH data, the data set containing positive case counts was retained. Of these data, daily reported case counts and case seven-day moving average were kept for analysis. These data were then subset to include only data from Clarke County collected during the surveillance period and combined with wastewater data.

@@

### Feature selection

Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where

### Data splitting

Compare relationship differences between year 1 and year 2

### Model training and fitting

Trained linear regression models on year 1 data using cross validation

Performance validation, metrics?

### Model testing

The trained model was then tested on the year 2 data

@@

```{r, echo=FALSE, message=FALSE}

readRDS(here("results/manuscript_table1.rds"))
```

@@
